---
title: "Course Project for the Statistical Inference Course as part of the Data Science Specialisation on Coursera"
author: "David Mullins"
output: pdf_document
---

## Overview

There are two tasks for this course project. The first is a simulation exercise using the exponential distribution. The second is a basic inferential data analysis.

```{r set_working_dir}
## Set the working directory for the analysis 
setwd("/home/david/Desktop/Coursera_DataScience/Statistical_inference_course_project")
```

## Part 1 - Simulation Exercise

```{r simulate_exp_data}
## Simulate data for an exponential distribution

### Set variables
num_obs <- 40
lambda <- 0.2
num_simulations <- 1000
rseed <- set.seed(1) ## Set the random seed for reproducibility

## Simulate the data from the exponential distribution and arrange as a matrix
exp_sim.matrix <- matrix( rexp(num_obs*num_simulations, rate = lambda), num_simulations)

mns <- apply(exp_sim.matrix, 1, mean) ## Calculate the mean for each sample
vars <- apply(exp_sim.matrix, 1, var) ## Calculate the variance for each sample

```

### Sample Mean versus Theoretical Mean

The theoretical mean for the exponential distribution is 1/lambda. This works out as 5. 

```{r theoretical_mean}
## theoretical mean
tmean <- 1/lambda
tmean
```

The sample mean for the 1,000 simulated distributions is 4.99.

```{r sample_mean}
## Summary stats for the collective means of each simulated distribution
summary(mns)
smean <- mean(mns)
smean
```

The below plot shows the distribution of the sample means of 1,000 simulated exponential distributions (with n=40 for each simulation). The theoretical mean is plotted in green at x=5. The sample mean is shown in red at x=4.99.

```{r mean_plot}
hist(
    x = mns,
    breaks = 20,
    freq = FALSE,
    main = "Means of 1000 simulated exponential distributions",
    xlab = "Sample Means"
)
abline(v = tmean, col = "green")
abline(v = smean, col =  "red")
```

### Sample Variance versus Theoretical Variance

The theoretical standard deviation for the exponential distribution is 1/lambda. This works out as 5. The variance is the standard deviation squared. Therefore, the theoretical variance of the mean is 25.

```{r theoretical_variance}
## theoretical variance
tvar <- (1/lambda)^2
tvar
```

The sample variance of the means for the 1,000 simulated distributions is 25.06.

```{r sample_variance}
## Summary stats for the collective means of each simulated distribution
summary(vars)
## sample variance
svar <- mean(vars)
svar
```

The below plot shows the distribution of the sample variance of 1,000 simulated exponential distributions (with n=40 for each simulation). The theoretical variance is plotted in green at x=25. The mean sample variance is shown in red at x=25.06.

```{r variance_plot}
hist(
    x = vars,
    breaks = 20,
    freq = FALSE,
    main = "Sample variances of 1000 simulated exponential distributions",
    xlab = "Sample Variances"
)
abline(v = tvar, col = "green")
abline(v = svar, col =  "red")
```

### Investigate Normality of the Distribution

The mean and median are approximately the same, which indicates normality. The histogram of the simulated data points seems to have half a bell-shaped curve. This also indicates normality. In the "Mean qqplot" the theoretical mean (green line) and the qqline of the sample means (red line) intersect at 0 on the x-axis. This fact, and that most of the data points lie on the red qqline is an indication of normality. A similar case is observed in the "Variances qqplot". It can be said that the distribution of the data is approximately normal.

```{r normality}
## If the mean and median are approximately the same, this is an indication of normality
mean(mns); median(mns)

## Plot a histogram of the distribution of the underlying data
hist(
    x = exp_sim.matrix,
    breaks = 30,
    freq = F,
    main = "Distribution of the simulated data points",
    xlab = "Exponential Data Values"    
)

## qqplot of the means distribution
qqnorm(
    mns,  
    main = "Means qqplot",
    xlab = "Theoretical Mean", 
    ylab = "Sample Means"
)
qqline(mns, col = "red")
abline(h = tmean, col = "green")
legend(
    "topleft", 
    lty = 1, 
    col = c("red", "green"), 
    legend = c("Sample quantiles line of mean", "theoretical mean")
)

## qqplot of the variances distribution
qqnorm(
    vars,  
    main = "Variances qqplot",
    xlab = "Theoretical Variance", 
    ylab = "Sample Variances"
)
qqline(vars, col = "red")
abline(h = tvar, col = "green")
legend(
    "topleft", 
    lty = 1, 
    col = c("red", "green"), 
    legend = c("Sample quantiles line of variance", "theoretical variance")
)
```